# Reproducible Research: Peer Assessment 1




## Loading and preprocessing the data

For loading the data, we assume that the data is available in the `data` directory at the same level as the markdown file. If it is not, the data should be [downloaded](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) and unzipped in the `data` directory. The unzipped file is named `activity.csv`


```{r echo=TRUE}
# Path to data
data_file_path <- "data/activity.csv"
if (!file.exists(data_file_path)){
    stop("Data file not available. Please place the data in `data/activity.csv`")
}
data <- read.csv(data_file_path)
data$date <- as.Date(data$date)
# Check how the data looks
head(data, n=5)
```

## What is mean total number of steps taken per day?

1. The histogram of the data can be plotted using the following piece of code

```{r histogram_steps_raw_data}
with(data, hist(steps))
```

2. The mean and meadian of the total number of steps can be calculated as

```{r}
mean_steps <- mean(data$steps, na.rm=TRUE)
mean_steps
median_steps <- median(data$steps, na.rm=TRUE)
median_steps
```

## What is the average daily activity pattern?

1. Time series plot of the 5-minute interval and the average number of steps taken, averaged across all days

```{r average_daily_activity_pattern}
library(data.table)
dt <- data.table(data)
mean_across_days <- dt[,list(mean=mean(steps, na.rm=TRUE)),by=interval]
with(mean_across_days, plot(interval, mean, type="l"))
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
mean_across_days[which.max(mean_across_days$mean),][["interval"]]
```

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
sum(is.na(data))
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
imputed_data <- data
na_data <- is.na(imputed_data)
na_intervals <- imputed_data$interval[na_data]
imputed_data[na_data] <- sapply(na_intervals, function(x) mean_across_days[interval==x,][["mean"]])
sum(is.na(imputed_data))
```

```{r histogram_steps_imputed_data}
with(imputed_data, hist(steps))
mean(imputed_data$steps)
median(imputed_data$steps)
```

As you can see, the mean and median of the raw and imputted data remains the same.
Now let's look at the number of steps:

```{r}
sum(data$steps, na.rm=TRUE)
sum(imputed_data$steps)
```

As expected, the number of steps for the imputed data is more than the raw data.

## Are there differences in activity patterns between weekdays and weekends?

```{r}
imputed_data$weekdays <- weekdays(imputed_data$date) %in% c('Saturday', 'Sunday')
imputed_data$weekfactors <- factor(imputed_data$weekdays, labels=c("weekday","weekend"), levels=c(FALSE, TRUE))
```

```{r weekday_weekend_pattern}
dt_imputed <- data.table(imputed_data)

data_to_plot <- dt_imputed[dt_imputed$weekfactors=="weekday",list(steps=mean(steps), weekfactors="weekday"), by=interval]

data_to_plot <- rbind(data_to_plot, dt_imputed[dt_imputed$weekfactors=="weekend",list(steps=mean(steps), weekfactors="weekend"), by=interval])

head(data_to_plot)

library(lattice)
xyplot(steps ~ interval | weekfactors, data=data_to_plot, type = "l", horizontal = TRUE, layout=c(1,2))
```
